---
title: "Quality Trimming - Anglerfish"
author: "Griffin Patel"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      fig.path = "../../figures/Lecture/01_QualityTrimming_Lecture/") # send any figure output to this folder 
```

# Goals of this file 

0. Load packages and prepare for fastq files. 
1. Use raw fastq files and generate quality plots to assess quality of raw reads.
2. Evaluate whether primers are present in the sequences.
3. Filter and trim out primers, bad sequences and bases from the sequencing files. 
4. Assess if filtering and trimming was adequate. If not, re-run with new parameters.
4. Write out fastq files with high quality sequences. 
5. Evaluate the quality from the filter and trim step and whether it is adequate. 


Output that we will create in this file: 

1. Quality plots to asses pre- and post-quality trimming. 
2. Written files that include the filtered and trimmed data. (Note that this will be written during the `filterAndTrim()` function.)

Why is it useful to set a seed? 

1. *Reproducibility*:
    - If you’re using random number generation (e.g., bootstrapping, simulations, machine learning models), setting a seed ensures that you (and others) can get the same results every time you run the code.
2. *Debugging & Collaboration*:
    - If your analysis involves randomness (e.g., shuffling datasets, training ML models), setting a seed makes it easier to debug and compare results across different runs or machines.
    - Colleagues can run your code and get identical results, making scientific collaboration more reliable.
3. *Benchmarking & Model Comparisons*:
    - In machine learning, setting a seed ensures that models trained with random splits of data or initial weights are comparable across different trials.
    - If you don’t set a seed, each run might give slightly different performance metrics, making it harder to evaluate true differences between models.
4. *Avoiding Accidental Bias in Random Sampling*:
    - When selecting random subsets of data (e.g., cross-validation splits), setting a seed ensures that all experiments are based on the same random selection, reducing unwanted variability.

```{r set-seed}
# Any number can be chosen 
set.seed(238428)
```

### Timing of Script
```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
start_time
```

# Load Libraries 
```{r load-libraries}
# Efficient package loading with pacman 
pacman::p_load(tidyverse, dada2, patchwork, DT,
               devtools, Biostrings, ShortRead,
               install = FALSE)
```

# Prepare Raw fastq Files 
```{r load-data}
# Set the raw fastq path to the raw sequencing files 
# Path to the fastq files 
raw_fastqs_path <- file.path(getwd(), "data", "00_raw_gzipped_fastqs")
raw_fastqs_path

# What files are in this path? Intuition Check 
head(list.files(raw_fastqs_path))

# How many files are there? 
length(list.files(raw_fastqs_path))

# Create vector of forward reads
forward_reads <- sort(list.files(raw_fastqs_path, pattern = "_1\\.fastq\\.gz$", full.names = TRUE))  

# Intuition Checks 
head(forward_reads)

# Create a vector of reverse reads 
reverse_reads <- sort(list.files(raw_fastqs_path, pattern = "_2\\.fastq\\.gz$", full.names = TRUE))

# Intuition Checks
head(reverse_reads)

# Intuition check #2: Need to have equal number of forward and reverse files 
stopifnot(length(reverse_reads) == length(forward_reads))

# Create sample names from forward reads
sample_names <- sub("_1\\.fastq\\.gz$", "", basename(forward_reads))

# Intuition Checks
head(sample_names)

# Intuition check #3: Forward and reverse sample names should match
stopifnot(identical(sample_names, sub("_2\\.fastq\\.gz$", "", basename(reverse_reads))))
```

# Assess the raw read quality 

```{r raw-read-quality}
# Randomly select 12 samples to assess 
random_samples <- sample(1:length(reverse_reads), size = 12)
random_samples

#Plot forward read quality 
forward_rawQual_plot12 <- 
  plotQualityProfile(forward_reads[random_samples]) + 
  labs(title = "Forward Reads: Raw Quality")

# Plot reverse read quality 
reverse_rawQual_plot12 <- 
  plotQualityProfile(reverse_reads[random_samples]) + 
  labs(title = "Reverse Reads: Raw Quality")

# Plot both plots together 
forward_rawQual_plot12 + reverse_rawQual_plot12
```

**REMEMBER:** If you have NextSeq data, your quality scores will be binned, instead of having the 40 integer phred scores. This is ok!


**Acceptable Phred Scores for ASVs with DADA2**

In DADA2, quality filtering is crucial for obtaining reliable biological sequences. Below are the acceptable **Phred score** thresholds for DADA2 preprocessing:

| **Q Score**   | **Error Probability**    | **DADA2 Interpretation**            |
|--------------|------------------------|------------------------------------|
| **40**        | 0.0001 (1 in 10,000) | Very high-quality , dream come true   | 
| **≥30**     | ≤ 0.001 (1 in 1,000) | Excellent quality, keep                |
| **25-30** | ~0.003 - 0.01        | Can be acceptable, but more important to trim low-quality tails |
| **20-25** | ~0.01 - 0.03         | Lower quality, only acceptable if read length is absolutely critical |
| **<20**    | > 0.01 (1 in 100)    | Poor quality, remove                    |
| **<10**     | >0.1 (1 in 10)       | Very unreliable, absolutely remove      |
---

Some summary take-aways:

- **30 is good** – Bases with Q30+ scores have a **99.9% accuracy rate**, making them highly reliable.  
- **25-30 can be acceptable**, but trimming is preferred to remove low-quality tails.  
- **20-25** should only be kept if read length is absolutely critical (*e.g.* to ensure forward & reverse read overlap)
- **<20 discarded** – Bases with Phred scores below 20 have an **error rate >1%** and should 100% be **trimmed from the dataset**.  

Remember, every base counts when we assign ASVs, they are called **sequence variants** for a reason! And, we want to be sure. 

## Prepare filtering and Trimming 

```{r prep-filter-Trim}
#Create vector of sample names 
sample_names <- sapply(strsplit(basename(forward_reads), "_"), `[`, 1)

# Create a test of sample names 
sample_names_R2 <- sapply(strsplit(basename(reverse_reads), "_"), `[`, 1)

# StopIfNot 
stopifnot(sample_names == sample_names_R2)

#Test 
head(sample_names)

# Provide path for the filtered reads 
filtered_fastqs_path <- "data/01_Quality_Trimming/filtered_fastqs"
filtered_fastqs_path

# Create 2 vectors: filtered_forward_reads & filtered_reverse_reads
filtered_forward_reads <- 
  file.path(filtered_fastqs_path, paste0(sample_names, "_R1_filtered.fastq.gz"))
filtered_reverse_reads <- 
   file.path(filtered_fastqs_path, paste0(sample_names, "_R2_filtered.fastq.gz"))

#intuition check 
length(filtered_forward_reads)
length(filtered_reverse_reads)
```

```{r raw-aggregate-plot, fig.width=5.5, fig.height=3.5}
# Aggregate all QC plots 
# Forward reads
forward_preQC_plot <- 
  plotQualityProfile(forward_reads, aggregate = TRUE) + 
  labs(title = "Forward Pre-QC")

# reverse reads
reverse_preQC_plot <- 
  plotQualityProfile(reverse_reads, aggregate = TRUE) + 
  labs(title = "Reverse Pre-QC")

# Now, let's put the two plots together
preQC_aggregate_plot <- 
  # Plot the forward and reverse together 
  forward_preQC_plot + reverse_preQC_plot
# Show the plot
preQC_aggregate_plot
```

<span style="color: red;">INTERPRETATION #1 of Pre-QC: Here, in this location of your analyses, please insert a description of the interpretation you draw regarding your plots. You must note anything important about the figure you've created about your data including any spikes in quality and also the general trend of the raw QC plots. An example is provided below.</span> 

My plots were generated with 2x250 bp sequencing on an Illumina MiSeq Sequencer, and show both the aggregated forward and reverse Pre-QC reads. There is a small drop in quality in the early bases, but none that are below 30, so there is no need to trim the front. Quality in both the forward and reverse sequences is stable and very high between 20 and 150 bp, and then steadily starts decreasing in quality from that point. The forward reads quality doesn't begin dropping below 30 on average until around 220-225bp, while the reverse reads reach a sub-30 quality average much sooner, around the 175bp mark. There are also very brief spikes in low quality in the reverse reads earlier, but never to an unusable quality threshold. It appears that both forward and reverse reads do not require trimming at the beginning, but the forward reads should have their later read tail trimmed to 220bp, and reverse reads to 200bp. This still puts the quality into the more questionable 25-30 range, but because quality is high elsewhere, in the interests of preserving as much read length, it will be used. This drop in quality as the reads go on is to be expected with illumina sequencing, as the error rate increases with later cycles. 

# 2. Evaluate Primer Search 


> Does your dataset include primers? 

Before we go ahead and filter and trim our reads, we need to confirm whether or not the raw seuqences include the primer sequences in them. Their presence depends on the type of libray prepration protocol that was used.

However, if the standard Illumina 2-step PCR library prep was used, **then primers will be included in the sequences and they will need to be trimmed out of the sequences later.**

Now, let's check for those primers! 

```{r search-primers-test}
# Search the sequences for primers to check if they are present
# Pull out the raw_fastq sequences from all of the raw sequencing files 
raw_fastqs <- readFastq(dirPath = raw_fastqs_path)

# The above code makes a complex object. Let's only grab the DNA sequences 
all_seqs <- sread(raw_fastqs)

f_primer <- "GTGYCAGCMGCCGCGGTAA"
r_primer <- "GGACTACNVGGGTWTCTAAT"

# Search the dataset 
# Check the complement of the primers
Fprimer_complement <- complement(DNAString(f_primer))
Fprimer_revComplement <- reverseComplement(DNAString(f_primer))

Rprimer_complement <- complement(DNAString(r_primer))
Rprimer_revComplement <- reverseComplement(DNAString(r_primer))

############ POSITIVE CONTROLS OF CODE TO MAKE SURE IT WORKS ############

# Use one real sequence from this dataset as a test sequence
test_DNAseq <- all_seqs[[1]]

# Create a positive control pattern from a real read in this dataset
real_pos_pattern <- as.character(subseq(test_DNAseq, start = 20, width = 15))

# Positive control intuition check
pos_control <-
  vcountPattern(pattern = real_pos_pattern,
                subject = all_seqs,
                fixed = FALSE)

paste("Positive control is found in",
      round(mean(pos_control > 0) * 100, digits = 2),
      "% of sequences")

# When there is a match, vmatchPattern can be used to locate the indices of the matches 
test_pos_control <- 
  vmatchPattern(pattern = real_pos_pattern, 
                subject = all_seqs, 
                fixed = FALSE)

# Take a look 
head(test_pos_control)

# Double check the degenerate nucleotides are ok within these functions, too 
pos_degen_control <-
  vcountPattern(pattern = "GTGYC", 
                subject = all_seqs, 
                fixed = FALSE)

paste("Positive control with a degenerate base is found in",
      round(mean(pos_degen_control > 0) * 100, digits = 2),
      "% of sequences") 

stopifnot(sum(pos_degen_control) > 0)
```

Now, let's specifically test for the **forward** primers: 

```{r forward-primers}
############ FORWARD PRIMERS ############ 
# Now, let's search the sequences for the primers
# Search first for forward primer 
Fprimer_vmatch <- 
  vcountPattern(pattern = f_primer, 
                subject = all_seqs, 
                fixed = FALSE); 
# Note: 
paste("The forward primer is found in", round(sum(Fprimer_vmatch)/length(all_seqs) * 100, digits = 2),
      "% of sequences") 


# Then double check with complement of forward primer 
Fprimer_complement_vmatch <- 
  vcountPattern(pattern = Fprimer_complement, 
                subject = all_seqs, 
                fixed = FALSE); 
# Note: 
paste("The complement of the forward primer is found in", round(sum(Fprimer_complement_vmatch)/length(all_seqs) * 100, digits = 2),
      "% of sequences") 
```

Great, now let's specifically test for the **reverse** primers: 

```{r reverse-primers}
############ REVERSE PRIMERS ############ 
# Now, let's search for the reverse primer 
Rprimer_vmatch <- 
  vcountPattern(pattern = r_primer, 
                subject = all_seqs, 
                fixed = FALSE); 
paste("The reverse primer is found in", round(sum(Rprimer_vmatch)/length(all_seqs) * 100, digits = 2),
      "% of sequences") 

# Then double check with complement of forward primer 
Rprimer_complement_vmatch <- 
  vcountPattern(pattern = Rprimer_complement, 
                subject = all_seqs, 
                fixed = FALSE); 
paste("The complement of reverse primer is found in", round(sum(Rprimer_complement_vmatch)/length(all_seqs) * 100, digits = 2),
      "% of sequences") 
```

# 3. Filter and Trim 

##  A. Prepare a placeholder for filtered reads 
```{r prep-filtered-sequences}
# Create vector of sample names from the filenames 
sample_names <- sapply(strsplit(basename(forward_reads), "_"), `[`,1) 

# Intuition Check 
head(sample_names)

# Place filtered reads into filtered_fastqs_path
filtered_fastqs_path <- "data/01_Quality_Trimming_Lecture/filtered_fastqs"

# Intuition Check 
filtered_fastqs_path

# create 2 vectors: filtered_forward_reads & filtered_reverse_reads
filtered_forward_reads <- 
  file.path(filtered_fastqs_path, paste0(sample_names, "_R1_filtered.fastq.gz"))

# Intuition Check 
length(filtered_forward_reads)

# reverse reads
filtered_reverse_reads <- 
  file.path(filtered_fastqs_path, paste0(sample_names, "_R2_filtered.fastq.gz"))

# Intuition Check 
head(filtered_reverse_reads)
```
## B. Actually Filter and Trim Reads
```{r filter-and-trim}
# Assign and write out filtered fastq files 
# Here, in this class dataset, the Kozich et al.(2013) AEM
      # Link to paper: https://doi.org/10.1128/AEM.01043-13
# Therefore, we do not need to trim the primers, because they were not sequenced
filtered_reads <- 
  filterAndTrim(fwd = forward_reads, filt = filtered_forward_reads,
              rev = reverse_reads, filt.rev = filtered_reverse_reads,
              maxN = 0, maxEE = c(2,2), 
              # Remove the first 8 bp in forward and 3bps in reverse reads
              # Note, this will be the place where primers will be removed! (see example below)
              trimLeft = c(8,3),
              # Keep the full forward & trim final 2 bases in the reverse with truncLen
              truncLen = c(250, 248), 
              truncQ = 2, rm.phix = TRUE, compress = TRUE, 
              # Please set threads to be 8-10 (we want to respect others using the server!)
              # Note that if TRUE, it will use ALL threads (making it hard for others to use the server)
              multithread = 40) 

##### Another example of a filterAndTrim Step 
# This dataset is from sequences generated by Cornell Sequencing Facility using  
# the typical Earth Microbiome protocol mentioned in their 16S SOP:
        # https://earthmicrobiome.org/protocols-and-standards/16s/
# Here, the library prep was the standard Illumina 2-step PCR preparation and following
# In this example, the primer sequences ARE in the sequences and must be trimmed using trimLeft
#
#filterAndTrim(forward_reads, filtered_forward_reads,
#              reverse_reads, filtered_reverse_reads,
#              truncLen = c(240,220), trimLeft = c(19,20),
#              maxN = 0, maxEE = c(1,1), truncQ = 2, 
#              rm.phix = TRUE, compress = TRUE, 
#              # Please set threads to be 8-10 (we want to respect others using the server!)
#              # Note that if TRUE, it will use ALL threads (making it hard for others to use the server)
#              # multithread = 10)
```
## C. Plot Filtered & Trimmed Read Quality of Subset of Samples

```{r filterTrim-quality-plots,  fig.width=12, fig.height=8}
# Plot the 12 random samples after QC
forward_filteredQual_plot_12 <- 
  plotQualityProfile(filtered_forward_reads[random_samples]) + 
  labs(title = "Trimmed Forward Read Quality")

reverse_filteredQual_plot_12 <- 
  plotQualityProfile(filtered_reverse_reads[random_samples]) + 
  labs(title = "Trimmed Reverse Read Quality")

# Put the two plots together 
forward_filteredQual_plot_12 + reverse_filteredQual_plot_12
```

## D. Plot Aggregated Filtered & Trimmed Plots of All Samples
```{r qc-aggregate-plot, fig.width=5.5, fig.height=3.5}
# Aggregate all QC plots 
# Forward reads
forward_postQC_plot <- 
  plotQualityProfile(filtered_forward_reads, aggregate = TRUE) + 
  labs(title = "Forward Post-QC")

# reverse reads
reverse_postQC_plot <- 
  plotQualityProfile(filtered_reverse_reads, aggregate = TRUE) + 
  labs(title = "Reverse Post-QC")

# Now, let's put the two plots together
postQC_aggregate_plot <- 
  # Plot the forward and reverse together 
  forward_postQC_plot + reverse_postQC_plot
# Show the plot
postQC_aggregate_plot
```

## E. Save Filtered Reads Data 

```{r save-filtered-reads-data}
# SAVE DATA FIRST 
save(filtered_reads, file = "data/01_Quality_Trimming/filtered_reads.RData")
```

# 4. Assess if filtering and trimming was adequate. If not, re-run with new parameters.

### A. Read Retention Post-QC

```{r filterTrim-stats, message = FALSE, fig.height = 2.5, fig.width=8}
# Make output into dataframe 
filtered_df <- as.data.frame(filtered_reads) %>%
  mutate(percent.retained = reads.out/reads.in)

# Intuition check
# Visualize it in table format 
DT::datatable(filtered_df)

# Let's calculate some statistics
read_stats_df <- 
  filtered_df %>%
  reframe(median_reads_in = median(reads.in),
          median_reads_out = median(reads.out),
          median_percent_retained = (median(reads.out)/median(reads.in)),
          max_percent_retained = max(reads.out/reads.in),
          min_percent_retained = min(reads.out/reads.in))

# Take a look at it!
read_stats_df

# Plot it 
numSeqs_QC_dotplot <-
  filtered_df %>%
  ggplot(aes(x = reads.in, y = reads.out)) + 
  geom_point(alpha = 0.5, size = 2) + 
  labs(x = "# of Raw Seqs", 
       y = "# of Seqs Retained") + 
  # Now let's add a 1:1 line for reference of keeping 100% of the reads
  geom_abline(slope=1, intercept = 0, color = "deeppink")

# Now, let's look at the number of reads retained in a histogram
numRetained_QC_histplot <- 
  filtered_df %>%
  ggplot(aes(x = reads.out)) + 
  geom_histogram() + 
  labs(x = "# of Seqs Retained", 
       y = "# of Samples") 

# Create a histogram of percent reads retained in a histogram
percSeqs_QC_histplot <- 
  filtered_df %>%
  ggplot(aes(x = percent.retained)) + 
  geom_histogram() + 
  labs(x = "% of Seqs Retained", 
       y = "# of Samples") + 
  # Set the scale to be between 0-1 (0-100%)
  scale_x_continuous(limits = c(0, 1))

# Now, let's put the plots together
numSeqs_QC_dotplot + numRetained_QC_histplot + percSeqs_QC_histplot + 
  plot_annotation(tag_levels = 'A')
```

## B. Visualize QC differences Pre- and Post-Filering & Trimming 
```{r pre-post-QC-plot, fig.width=6, fig.height=5.5}
# Plot the pre and post together in one plot
preQC_aggregate_plot / postQC_aggregate_plot
```

# Check Render Time
```{r stop-time}
# Take the time now that we are at the end of the script
end_time <- Sys.time()
end_time 

# Echo the elapsed time
elapsed_time <- round((end_time - start_time), 3)
elapsed_time
```

# Session Information 
```{r session-info}
# Ensure reproducibility 
devtools::session_info()
```

