---
title: "Assigning ASVs with DADA2 - Anglerfish"
author: "Griffin Patel"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center", 
                      fig.path = "../figures/02_AssignASVs")
```

# Goals 

1. Infer an error model for the filtered sequences, separately on forward and reverse reads.
2. Assign ASVs on both forward and reverse reads separately by applying the error model.
3. Merge forward and reverse ASVs into contiguous ASVs.
4. Generate the first draft of the ASV count table.
5. Quality trimming of ASV lengths.
6. Remove chimeras. 
7. Assign taxonomy with the Silva database. 
8. Write out relevant files: `asv_table`, `asvs_fasta`, `tax_table`, and `sample_data`.

## Input 

1. Filtered fastq files generated from `01_QualityTrimmingAnglerfish.Rmd`.
2. Sample name vector.

## Output 

1. ASV Count Table: `asv_table` (with and without sequence names)
2. ASV fasta file: `asvs_fasta`
3. Taxonomy table: `tax_table`
4. Sample information: `sample_data`

# Set up the Environment 

## Set Seed & Threads

```{r set-seed}
# Set the seed for reproducibility
set.seed(238428)

# Set the number of threads
n_threads <- 20
```

## Timing of this script

```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```

## Load Packages 

```{r load-packages}
# Install packages if needed
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

if (!requireNamespace("dada2", quietly = TRUE)) {
  BiocManager::install("dada2", ask = FALSE, update = FALSE)
}

if (!requireNamespace("phyloseq", quietly = TRUE)) {
  BiocManager::install("phyloseq", ask = FALSE, update = FALSE)
}

if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}

if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}

if (!requireNamespace("patchwork", quietly = TRUE)) {
  install.packages("patchwork")
}

if (!requireNamespace("DT", quietly = TRUE)) {
  install.packages("DT")
}

library(tidyverse)
library(devtools)
library(dada2)
library(phyloseq)
library(patchwork)
library(DT)
```

## Load Filtered Fastq Files 

```{r load-filtered-fastqs}
# Place filtered seq files into a variable 
filtered_fastqs_path <- "/local/workdir/glp56/Anglerfish_16S/data/01_Quality_Trimming_Lecture/filtered_fastqs"

# Intuition check:
filtered_fastqs_path

# Create forward vector 
filtered_forward_reads <- 
  sort(list.files(filtered_fastqs_path, 
                  pattern = "R1_filtered.fastq.gz$", 
                  full.names = TRUE))

# Check 
filtered_forward_reads[1:5]

# Create reverse vector 
filtered_reverse_reads <- 
  sort(list.files(filtered_fastqs_path, 
                  pattern = "R2_filtered.fastq.gz$", 
                  full.names = TRUE))

# Check 
filtered_reverse_reads[1:5]

# Intuition checks
length(filtered_forward_reads)
length(filtered_reverse_reads)

stopifnot(length(filtered_forward_reads) > 0)
stopifnot(length(filtered_forward_reads) == length(filtered_reverse_reads))
```

## Sample Names 

```{r sample-names}
# Create vector of sample names from the filenames 
sample_names <- sapply(strsplit(basename(filtered_forward_reads), "_"), `[`, 1) 

# Intuition check 
head(sample_names)

# Confirm reverse filenames match
stopifnot(
  identical(
    sample_names,
    sapply(strsplit(basename(filtered_reverse_reads), "_"), `[`, 1)
  )
)
```

# Error Modelling 

## Learn the Errors

### MiSeq Runs: 40 Phred Scores

```{r learn-errors-MiSeq, fig.width=12, fig.height=8}
# Forward reads 
error_forward_reads <- 
  learnErrors(filtered_forward_reads, multithread = n_threads)

# Forward error plot 
forward_error_plot <- 
  plotErrors(error_forward_reads, nominalQ = TRUE) + 
  labs(title = "Forward Reads: Error Model")

# Reverse reads 
error_reverse_reads <- 
  learnErrors(filtered_reverse_reads, multithread = n_threads)

# Reverse error plot 
reverse_error_plot <- 
  plotErrors(error_reverse_reads, nominalQ = TRUE) + 
  labs(title = "Reverse Reads: Error Model")

# Look at the plots together 
forward_error_plot + reverse_error_plot
```

# Infer ASVs

**An important note:** What makes DADA2 so nice is that it infers ASVs separately on the forward and the reverse reads and then merges them afterwards. This is quite a different approach from how OTUs are identified in Mothur and also from UCHIME, oligotyping, and other OTU, MED, and ASV approaches.

```{r infer-ASVs}
# Infer ASVs on the forward sequences
dada_forward <- 
  dada(filtered_forward_reads, 
       err = error_forward_reads,
       multithread = n_threads) 

# Take a look at the data
# What type of data structure is it? 
typeof(dada_forward) # It's a list 
length(dada_forward) # How big is it? One per sample!

# What does it look like for each sample?  
dada_forward[[1]]

# Reverse ASVs
dada_reverse <- 
  dada(filtered_reverse_reads,
       err = error_reverse_reads,
       multithread = n_threads)

# Check data 
dada_reverse[[1]]
```

# Merge Forward and Reverse ASVs 

Now, that we have identified our ASVs separately on both the forward and the reverse reads, let's **merge** them together into contiguous (*aka* "contigs") ASVs.

```{r merge-ASVs}
merged_ASVs <- 
  mergePairs(dada_forward, filtered_forward_reads,
             dada_reverse, filtered_reverse_reads,
             verbose = TRUE)

# Evaluate the data output 
typeof(merged_ASVs) # A list
length(merged_ASVs) # Length of the number of samples!
head(names(merged_ASVs)) # Here, we can access our current sample names

# Inspect further for each sample
glimpse(merged_ASVs[[1]])
```

# Create Raw ASV Count Table 

```{r raw-ASV-count-table}
# Raw ASV
raw_ASV_table <- makeSequenceTable(merged_ASVs)

# Intuition Check: Type and dimensions of the data
dim(raw_ASV_table)
typeof(raw_ASV_table)
class(raw_ASV_table)

# Write out raw_ASV_table 
dir.create("data/02_AssignASVs_Lecture", recursive = TRUE, showWarnings = FALSE)

write.table(raw_ASV_table, 
            file = "data/02_AssignASVs_Lecture/raw_ASV_counts.tsv",
            sep = "\t", quote = FALSE, col.names = NA)
```

# Assess ASV Quality 

## What are the current ASV lengths? 

First, let's inspect the distribution of the ASV sequence lengths across all of the ASVs that we have inferred. First, we need to consider: **What is the expected length of the ASVs?** 

1. *What is the sequencing strategy?*  
    - paired-end 250 bp (2x250 bp) Illumina MiSeq sequencing was used.  
    - Specifically, the V4 hypervariable region of the 16S rRNA gene was targeted for sequencing with the 515F and 806R primers.  

2. *What's the total length of our starting amplicons?*  
    - Our primers are named after the location they bind to on the 16S gene. So:  
    - `806 - 515 = 291`  
    - The total length of the starting amplicon is 291 base pairs, with primers.  

3. *What is the ASV length without primers?*  
    - The 515F (`GTGYCAGCMGCCGCGGTAA`) primer is 19 base pairs.  
    - The 806R (`GGACTACNVGGGTWTCTAAT`) primer is 20 base pairs.  
    - `291 - 19 - 20 = 252`  
    - The expected ASV length without primers is **252 base pairs**.  
    - Importantly, primers were not detected in this dataset, so 252 bp is the biologically expected ASV length before considering read trimming.

4. *What is the length expected after the `filterAndTrim()` step?*  
    - In the quality trimming step, the forward reads were truncated to 220 bp and the reverse reads were truncated to 200 bp, with no trimming from the start of either read.  
    - Because the full V4 amplicon is only about 252 bp, the merged ASV should still reconstruct the full amplicon length if overlap is sufficient.  
    - Therefore, the expected merged ASV length is still approximately **252 bp**.  

5. *What is the expected overlap between our forward and reverse reads?*  
    - The total retained read length after filtering is `220 + 200 = 420 bp`.  
    - If the true amplicon length is about 252 bp, then the expected overlap is `420 - 252 = 168 bp`.  
    - This is a very large overlap and should be more than sufficient for accurate merging.

<span style="color: red;">INTERPRETATION #1: My project data was generated using paired-end 2x250 bp Illumina MiSeq sequencing targeting the V4 region of the 16S rRNA gene with the 515F and 806R primers. The total starting amplicon length with primers is 291 bp. The forward primer is 19 bp long and the reverse primer is 20 bp long, so the expected amplicon length without primers is 252 bp. In my dataset, the primers were not detected in the reads, so 252 bp is the biologically expected ASV length before considering any quality trimming. In the filterAndTrim step, I did not trim any bases from the beginning of the reads, and I truncated the forward reads to 220 bp and the reverse reads to 200 bp. Even after this trimming, the merged ASV is still expected to reconstruct the full ~252 bp V4 amplicon, because the combined retained read length is 420 bp. This gives an expected overlap of about 168 bp between the forward and reverse reads, which is 66.7% of the amplicon length and is more than enough for accurate merging. Therefore, the hypothesized ASV length for my dataset is about 252 bp.
### ASV Length Stats 

```{r assess-ASV-length}
# Calculate summary stats
# Longest ASV?
maxLength_ASV <- max(nchar(getSequences(raw_ASV_table)))

# Shortest ASV?
minLength_ASV <- min(nchar(getSequences(raw_ASV_table))) 

# Mean ASV length?
meanLength_ASV <- mean(nchar(getSequences(raw_ASV_table))) 

# Median ASV length?
medianLength_ASV <- median(nchar(getSequences(raw_ASV_table))) 

# Create a table to inspect the distribution of sequence lengths of all ASVs in dataset 
table(nchar(getSequences(raw_ASV_table)))
```

Here's some of the stats that we've calculated:

- Max ASV length: `r maxLength_ASV` base pairs
- Min ASV length: `r minLength_ASV` base pairs
- Mean ASV length: `r meanLength_ASV` base pairs
- Median ASV length: `r medianLength_ASV` base pairs

### ASV Length Plot 

Now, we will plot these ASV lengths to give us a better idea: 

```{r ASV-lengths-plot}
# Inspect the distribution of sequence lengths of all ASVs in data set 
plot_ASVLength_raw <- 
  data.frame(Seq_Length = nchar(getSequences(raw_ASV_table))) %>%
  ggplot(aes(x = Seq_Length)) + 
  geom_histogram() + 
  # include the x-axis scales
  scale_x_continuous(limits = c(0, maxLength_ASV + 5)) + 
  labs(y = "Number of ASVs", x = "Raw ASV Length (bps)")

# Show the plot
plot_ASVLength_raw
```

<span style="color: red;">INTERPRETATION #2: The ASV length data largely matches my hypothesis from Interpretation 1, which predicted an expected ASV length of approximately 252 bp. The observed distribution shows a strong peak at 253 bp, with a smaller peak at 254 bp, which is very close to the expected value and within normal biological and sequencing variation. This indicates that the majority of ASVs were correctly reconstructed during merging. However, the presence of additional ASVs at shorter and longer lengths does not match the hypothesis and likely represents sequencing artifacts such as truncated reads, chimeras, or read-through into adapter sequences. Overall, the dominant peak supports the hypothesis, while the low-frequency off-target lengths indicate noise that should be removed.

***

<span style="color: red;">INTERPRETATION #3: Based on the ASV length distribution, I will retain only sequences that match the dominant biological length of 253 bp and remove all other ASVs. The strong peak at 253 bp indicates that this is the correct reconstructed length of the V4 amplicon in this dataset, while the other lengths represent noise or artifacts. Keeping only the dominant length prevents misformed ASVs from affecting diversity estimates and taxonomic assignment. This filtering step improves the overall quality and reliability of the dataset.

# Trim ASV lengths 

Now that we are convinced that the dominant biological ASV length in this dataset is 253 bp, we will trim the ASV table to retain only ASVs of that length. This removes lower-frequency off-target lengths that are likely artifacts.

```{r trim-ASVs}
# Subset only ASVs that are 253 bp long 
raw_ASV_table_trimmed <- 
  raw_ASV_table[, nchar(colnames(raw_ASV_table)) == 253]

# Inspect the distribution of sequence lengths of all ASVs in dataset 
table(nchar(getSequences(raw_ASV_table_trimmed)))

# What proportion of total ASV sequences are left in the data? 
percRetained_Trimmed <- sum(raw_ASV_table_trimmed) / sum(raw_ASV_table)
percRetained_Trimmed

# Inspect the distribution of sequence lengths of all ASVs in dataset 
# AFTER TRIM
plot_ASVLength_trimmed <- 
  data.frame(Seq_Length = nchar(getSequences(raw_ASV_table_trimmed))) %>%
  ggplot(aes(x = Seq_Length)) + 
  geom_histogram() + 
  scale_x_continuous(limits = c(0, maxLength_ASV + 5)) + 
  labs(y = "Number of ASVs", x = "Trimmed ASV Length (bps)")

# Show the plot 
plot_ASVLength_trimmed
```

# Remove Chimeras

Chimeras are artificial sequences formed during PCR when incomplete extension products prime different templates in later cycles. These do not represent real biological sequences and must be removed before downstream analysis.

```{r rm-chimeras, fig.width=3.5, fig.height=3}
# Remove the chimeras in the trimmed ASV table
noChimeras_ASV_table <- 
  removeBimeraDenovo(raw_ASV_table_trimmed, 
                     method = "consensus", 
                     multithread = n_threads, 
                     verbose = TRUE)

# Check the dimensions
dim(noChimeras_ASV_table)

# What proportion is left of the sequences? 
# Chimera removal compared to trimming 
percRetained_chimerasTrimmed <- sum(noChimeras_ASV_table) / sum(raw_ASV_table_trimmed)

# Chimera removal compared to raw  
percRetained_chimerasRaw <- sum(noChimeras_ASV_table) / sum(raw_ASV_table)

# Plot it 
plot_ASVLength_NoChimeras <- 
  data.frame(Seq_Length_NoChim = nchar(getSequences(noChimeras_ASV_table))) %>%
  ggplot(aes(x = Seq_Length_NoChim)) + 
  geom_histogram() + 
  scale_x_continuous(limits = c(0, maxLength_ASV + 5)) + 
  labs(y = "Number of ASVs \n (Post-Chimera Removal)", 
       x = "ASV Length (bps)")

# Show the plot
plot_ASVLength_NoChimeras
```

INTERPRETATION #4: The proportion of raw ASVs retained after trimming was 0.965, indicating that the vast majority of sequences were at the expected biological length of 253 bp and that only a small amount of off-target noise was removed. After chimera removal, the proportion of trimmed ASVs retained was 0.986, showing that very few sequences were identified as chimeric. Overall, the proportion of total merged ASVs retained after both trimming and chimera removal was 0.952, meaning that most of the data was preserved through both steps. These results suggest that the ASV length trimming was appropriate, as it removed low-frequency artifacts while retaining the dominant biological signal and maintaining a high proportion of usable sequences.

### Plot ASV Lengths 

```{r plot-ASVLengths, fig.height=2.5, fig.width=7}
plot_ASVLength_raw + plot_ASVLength_trimmed + plot_ASVLength_NoChimeras + 
  plot_annotation(tag_levels = 'A')
```

# Assign Taxonomy 

Now, we will assign taxonomy to our ASVs using the Silva reference database.

```{r assign-tax}
# Assign taxonomy up to genus level
taxa_train <- 
  assignTaxonomy(
    noChimeras_ASV_table,
    refFasta = "/local/workdir/in_class_data/taxonomy/silva_nr99_v138.2_toGenus_trainset.fa.gz",
    multithread = n_threads
  )

# Add species-level assignment
taxa_addSpecies <- 
  addSpecies(
    taxa_train,
    refFasta = "/local/workdir/in_class_data/taxonomy/silva_v138.2_assignSpecies.fa.gz"
  )

# Inspect taxonomy
taxa_print <- taxa_addSpecies
rownames(taxa_print) <- NULL
head(taxa_print)
```

# Track the read counts

Here, we will look at the number of reads that were lost in the filtering, denoising, merging, and chimera removal.

```{r track-reads, fig.width=6, fig.height=4}
# A little function to identify number seqs 
getN <- function(x) sum(getUniques(x))

# Make the table to track the seqs 
track <- cbind(
  sapply(dada_forward, getN),
  sapply(dada_reverse, getN),
  sapply(merged_ASVs, getN),
  rowSums(noChimeras_ASV_table)
)

# Update column names
colnames(track) <- c("denoisedF", "denoisedR", "merged", "nochim")
rownames(track) <- row.names(noChimeras_ASV_table)

# Generate a dataframe to track the reads through our DADA2 pipeline
track_counts_df <- 
  track %>%
  as.data.frame() %>%
  rownames_to_column(var = "sample_names")

# Intuition check
stopifnot(track_counts_df$sample_names == row.names(noChimeras_ASV_table))

# Add a new column with the number of ASVs
track_counts_df <- 
  track_counts_df %>%
  mutate(num_ASVs = rowSums(noChimeras_ASV_table > 1))

# Visualize it in table format 
DT::datatable(track_counts_df)

# Plot it
track_counts_df %>%
  pivot_longer(denoisedF:nochim, names_to = "read_type", values_to = "num_reads") %>%
  mutate(read_type = fct_relevel(read_type, "denoisedF", "denoisedR", "merged", "nochim")) %>%
  ggplot(aes(x = read_type, y = num_reads, fill = read_type)) + 
  geom_line(aes(group = sample_names), color = "grey") + 
  geom_point(shape = 21, size = 3, alpha = 0.8) + 
  scale_fill_brewer(palette = "Spectral") + 
  labs(x = "Filtering Step", y = "Number of Sequences") + 
  theme_bw()
```

# Plot read depth and ASV richness

```{r numSeqsASV-plot, fig.height=2.5, fig.width=7}
plot_ReadDepth <- 
  track_counts_df %>%
  ggplot(aes(x = nochim)) + 
  geom_histogram() + 
  labs(x = "Total # of Sequences", y = "# of Samples") + 
  theme_bw()

plot_ASVRichness <- 
  track_counts_df %>%
  ggplot(aes(x = num_ASVs)) + 
  geom_histogram() + 
  labs(x = "Total # of ASVs", y = "# of Samples") + 
  theme_bw()

plot_ReadDepth_ASVRichness <- 
  track_counts_df %>%
  ggplot(aes(x = nochim, y = num_ASVs)) + 
  geom_point() + 
  labs(x = "Total # of Sequences", y = "# of ASVs") + 
  theme_bw()

plot_ReadDepth + plot_ASVRichness + plot_ReadDepth_ASVRichness + 
  plot_annotation(tag_levels = 'A')
```

<span style="color: red;">INTERPRETATION #5: Plot A shows that most samples have similar sequencing depth, generally clustering below ~100,000 reads, with a few samples having much higher read counts. This suggests overall consistent sequencing, but with some variability. There is an outlier of around 18-19 samples with frewer sequences, so this may suggest some grouped lesser associated taxa.

Plot B shows that most samples have relatively low to moderate ASV richness (most below ~300 ASVs), while a small number of samples have much higher richness, indicating potential outliers or samples with unusually high diversity.

Plot C shows a weak positive relationship between sequencing depth and ASV richness. While samples with more reads tend to have slightly more ASVs, the relationship is not that strong, and several samples with high read counts do not have proportionally higher ASV richness. These few sequences with extremely high ASV counts suggests a group or sepcies of dominant taxa. 

Becuase this paper delt with an endosymbiont of anglerfish, there is a natural outlier we would exepect to be dominant, which can explain some of the points in these plots. Overall, the dataset appears generally consistent, but the presence of outliers in both sequencing depth and ASV richness suggests that some samples should be examined more closely in downstream analyses.
# End Time

```{r stop-time}
end_time <- Sys.time()
end_time

elapsed_time <- round((end_time - start_time), 3)
elapsed_time
```



