---
title: "Pre-processing with phyloseq - Anglerfish"
author: "Griffin Patel"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      fig.path = "../figures/Lecture/03_PreProcessing_Lecture/")
```

# Goals

The goal with this file is to pre-process any other contaminants and unexpected ASVs from our sample. We will need to: 

1. Remove any contaminating ASVs that are **chloroplasts**. 
2. Remove ASVs that are **mitochondria**. 
3. Evaluate any ASVs from the **negative controls**. Then, remove negative controls. 
4. Evaluate the mock community or **positive control** to learn the accuracy of sequencing. 
5. Check for **reverse complements**. 
6. Check the **sequencing depth** of samples. Remove samples that obviously have too few reads. 
7. Add the metadata for this dataset only if needed. 
8. Write a `preprocessed_physeq` to be used in the next step of our workflow. 

## Input 

Here, we will start with the phyloseq object that we created in 02_AssignASVsAnglerfish.Rmd

1. **Raw phyloseq object**: We will need `data/02_AssignASVs_Lecture/raw_physeq.RData`. 

## Output

1. We will write out a complex R data object known as an **S4** object, which in this case is known as a phyloseq output. We will store this in `data/03_PreProcessing/preprocessed_physeq.RData`.  

# Set up the Environment 

## Set the seeds and threads 

```{r seed-threads}
# Set the seed for reproducibility
set.seed(238428)
```

## Timing of this script

```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```

## Load packages  
```{r load-libraries}
# Load packages 
pacman::p_load(devtools, phyloseq, Biostrings, tidyverse, dada2, patchwork,
               ggpubr, rstatix, install = FALSE)
```

# Load Phyloseq Data

```{r phyloseq-handoff}
# Load the data 
load("data/02_AssignASVs_Lecture/raw_physeq.RData")

# Look at the data 
raw_physeq
```

# Clean up the data

**Chloroplasts** and **mitochondria** need to be removed from 16S datasets because they contain their own 16S rRNA genes, which are evolutionarily derived from bacteria. These organelles can be unintentionally amplified during PCR, especially in host-associated samples, leading to misleading results by inflating microbial diversity.

Removing these sequences ensures the analysis reflects true bacterial and archaeal diversity in the sample.

## 1. Remove Chloroplasts

```{r rm-chloros}
# Make new physeq without chloroplasts
noChloros_physeq <- 
  raw_physeq %>% 
  # rm chloroplasts, which are Order
  subset_taxa(Order != "Chloroplast" | is.na(Order))
  
# How many taxa were chloroplasts? 
num_chloro_ASVs <- ntaxa(raw_physeq) - ntaxa(noChloros_physeq)
num_chloro_ASVs
```

There were `r num_chloro_ASVs` ASVs that were chloroplasts.

## 2. Remove Mitochondria  

```{r rm-mitos}
# remove mitochondria 
noChlorosMitos_physeq <- 
  noChloros_physeq %>%
  subset_taxa(Family != "Mitochondria" | is.na(Family))

# How many mitochondrial ASVs? 
num_mito_ASVs <- ntaxa(noChloros_physeq) - ntaxa(noChlorosMitos_physeq)
num_mito_ASVs

# Let's take a quick look 
noChlorosMitos_physeq
```

There were `r num_mito_ASVs` ASVs that were mitochondria in the data set.

<span style="color: red;">INTERPRETATION #1: How many chloroplast and/or mitochondrial ASVs in your dataset? Does this seem reasonable?: A total of 227 ASVs were removed as chloroplast and/or mitochondrial sequences, because the dataset decreased from 8788 taxa in the raw phyloseq object to 8561 taxa after chloroplast and mitochondrial filtering. This seems reasonable, since some mitochondria and chloroplast sequences would likely get sequenced by accident, but it shouldn't be a significant amount. Removing these ASVs should improve the accuracy of the downstream analysis. Around 97% of sequences remained after chloroplast and mitochondrial sequences were removed, which seems like a fair amount of error, but still good to have removed.</span>



```{r rm-mitos-chloros}
# How many total asvs were removed from chloros and mitos? 
ntaxa(raw_physeq) - ntaxa(noChlorosMitos_physeq)

# proportion of asvs kept? 
ntaxa(noChlorosMitos_physeq)/ntaxa(raw_physeq)
```

# Evaulate and remove the control samples 

Now, let's take a look at the negative controls and then make a decision about whether or not to remove the ASVs that we found in our controls.

## 3. Negative Controls

```{r detect-neg-controls}
# Pull metadata
metadata_df <- 
  noChlorosMitos_physeq %>%
  sample_data() %>% 
  data.frame()

# Make sure names exists
if(!("names" %in% colnames(metadata_df))){
  metadata_df$names <- rownames(metadata_df)
}

# Detect likely negative controls
neg_control_samples <- unique(c(
  rownames(metadata_df)[grepl("control|blank|ntc|negative|water", rownames(metadata_df), ignore.case = TRUE)],
  unlist(lapply(metadata_df, function(x) {
    rownames(metadata_df)[grepl("control|blank|ntc|negative|water", as.character(x), ignore.case = TRUE)]
  }))
))

neg_control_samples <- neg_control_samples[!is.na(neg_control_samples)]
neg_control_samples
length(neg_control_samples)
```

### 3a. Evaluate Negative Control ASVs in Samples vs Controls

```{r neg-control-physeq, fig.width=7, fig.height=3}
if(length(neg_control_samples) > 0){

  # Create a phyloseq object with only negative controls 
  neg_control_physeq <- 
    noChlorosMitos_physeq %>%
    subset_samples(., names %in% neg_control_samples) %>%
    # force remove ASVs with 0 counts 
    prune_taxa(taxa_sums(.) > 0, .)

  # Let's take a look
  neg_control_physeq

  # vector of negative control asvs
  neg_control_ASVs <- taxa_names(neg_control_physeq)

  # Visualize raw abundance values of control ASVs
  neg_control_physeq %>%
    plot_bar(., "ASV", fill = "ASV")

} else {
  message("No negative controls detected in this dataset. Skipping negative control evaluation.")
}
```

### 3b. Evaluate Negative Control ASVs in Samples vs Controls

```{r negControl-ASVs-SamplesVSControls, fig.width=5, fig.height=4}
if(length(neg_control_samples) > 0){

  # Add Sample_or_Control if missing
  metadata_df2 <- 
    noChlorosMitos_physeq %>%
    sample_data() %>%
    data.frame()

  if(!("Sample_or_Control" %in% colnames(metadata_df2))){
    metadata_df2$Sample_or_Control <- ifelse(metadata_df2$names %in% neg_control_samples, "Control", "Sample")
    rownames(metadata_df2) <- metadata_df2$names
    sample_data(noChlorosMitos_physeq) <- metadata_df2
  }

  # Make new phyloseq object with only the control ASVs and ALL samples 
  neg_controlASV_physeq <- 
    noChlorosMitos_physeq %>%
    subset_taxa(., ASV %in% neg_control_ASVs)

  # Inspect it 
  neg_controlASV_physeq

  # Plot Abundances of neg control ASVs in samples versus controls 
  neg_controlASV_physeq %>%
    plot_bar(., "ASV", fill = "ASV", facet_grid = Sample_or_Control~.)

} else {
  message("No negative controls detected in this dataset. Skipping control comparison.")
}
```

```{r negControl-byASV, fig.width=8, fig.height=7}
if(length(neg_control_samples) > 0){

  # Calculate the ASV relative abundance 
  asv_negControls_df <- 
    neg_controlASV_physeq %>%
    # agglomerate at the ASV level 
    tax_glom(taxrank = "ASV") %>% 
    # Melt to a long format 
    psmelt()

  # Check the taxonomy of each ASV
  asv_negControls_df %>%
    dplyr::select(Kingdom:ASV) %>%
    distinct()

  dim(asv_negControls_df)

  # Statistically test
  negControl_ASV_statTest <- 
    asv_negControls_df %>%
    group_by(ASV) %>%
    wilcox_test(Abundance ~ Sample_or_Control) %>%
    adjust_pvalue(method = "fdr") %>%
    add_significance() %>%
    add_xy_position(x = "Sample_or_Control", fun = "max", scales = "free") %>%
    mutate(y.position = y.position * 0.82)

  ggboxplot(
    asv_negControls_df, 
    x = "Sample_or_Control", y = "Abundance", 
    color = "Sample_or_Control",
    facet.by = "ASV", scales = "free_y", add = "jitter") + 
    stat_pvalue_manual(negControl_ASV_statTest) + 
    scale_color_manual(values = c("firebrick3", "cornflowerblue")) + 
    scale_fill_manual(values = c("firebrick3", "cornflowerblue")) + 
    theme_bw() + 
    theme(axis.text.x = element_blank(), axis.title.x = element_blank(),
          legend.title = element_blank(), legend.position = c(0.85, 0.07)) 

} else {
  message("No negative controls detected in this dataset. Skipping ASV-level control statistics.")
}
```

### 3c. Remove any ASVs in the negative controls. 

```{r rm-negControl-ASVs}
if(length(neg_control_samples) > 0){

  # Remove control samples only
  noControlsChlorosMitos_physeq <- 
    noChlorosMitos_physeq %>%
    subset_samples(., !(names %in% neg_control_samples))

  # Inspect no controls object 
  noControlsChlorosMitos_physeq

  # After filtering, how many ASVs were removed? 
  total_ASVs_rm <- ntaxa(noChlorosMitos_physeq) - ntaxa(noControlsChlorosMitos_physeq)
  total_ASVs_rm

} else {

  noControlsChlorosMitos_physeq <- noChlorosMitos_physeq
  total_ASVs_rm <- 0
}
```

<span style="color: red;">INTERPRETATION #2: Did your dataset have any negative controls? If so, how many ASVs did you remove? Does that seem reasonable to you? : My dataset did not appear to contain any negative controls. The code detected no negative control samples, so no control samples or ASVs were removed at this step.</span>

# 4. Positive Controls 

### What is a positive control? 

A **positive control** is a sample that should produce a known result, which allows researchers to verify accuracy, detect bias or contamination, and troubleshoot technical problems.

```{r detect-pos-controls}
# Detect likely mock or positive controls
metadata_df3 <- 
  noControlsChlorosMitos_physeq %>%
  sample_data() %>% 
  data.frame()

if(!("names" %in% colnames(metadata_df3))){
  metadata_df3$names <- rownames(metadata_df3)
}

mock_samples <- unique(c(
  rownames(metadata_df3)[grepl("mock|zymo|positive|pos", rownames(metadata_df3), ignore.case = TRUE)],
  unlist(lapply(metadata_df3, function(x) {
    rownames(metadata_df3)[grepl("mock|zymo|positive|pos", as.character(x), ignore.case = TRUE)]
  }))
))

mock_samples <- mock_samples[!is.na(mock_samples)]
mock_samples
length(mock_samples)
```

```{r eval-accuracy}
if(length(mock_samples) > 0){

  # Make mock physeq object 
  mock_physeq <- 
    noControlsChlorosMitos_physeq %>%
    subset_samples(., names %in% mock_samples) %>%
    prune_taxa(taxa_sums(.) > 0, .)

  # Inspect it 
  mock_physeq
  tax_table(mock_physeq)

  # write us a message summarizing this
  cat("DADA2 inferred", ntaxa(mock_physeq), "ASVs present in the mock community.")

} else {
  message("No positive control or mock community detected in this dataset. Skipping mock evaluation.")
}
```

<span style="color: red;">INTERPRETATION #3: Did your dataset have a mock community? If so, what can you say about the accuracy of your sequencing run? : My dataset did not appear to contain a mock community or other positive control. The code detected no mock or positive control samples, so this step was skipped. Because there was no mock community in the dataset, I cannot directly evaluate sequencing accuracy here using a known reference community.</span>

## Remove the mock community and it's ASVs

```{r rm-mock-asvs}
if(length(mock_samples) > 0){

  # Make a new phyloseq object without mock
  noControlsMock_ChlorosMitos_physeq <- 
    noControlsChlorosMitos_physeq %>%
    subset_samples(., !(names %in% mock_samples))

  # Inspect
  noControlsMock_ChlorosMitos_physeq

} else {

  noControlsMock_ChlorosMitos_physeq <- noControlsChlorosMitos_physeq
}
```

# 5. Reverse Complements 

## What is a reverse complement? 

DNA is double-stranded, and each strand is made up of complementary bases: A pairs with T and C pairs with G.

```{r reverse-complements}
# Pull out Taxa Table
tax_table_df <- 
  noControlsMock_ChlorosMitos_physeq %>%
  tax_table() %>% 
  data.frame()

# Add ASV column if needed
if(!("ASV" %in% colnames(tax_table_df))){
  tax_table_df$ASV <- rownames(tax_table_df)
}

# Grab ASV Sequences and Names
asv_names <- tax_table_df$ASV
asv_seqs <- rownames(tax_table_df)

# Create empty comparison matrix
asv_rev_comp <- matrix(nrow = length(asv_seqs), ncol = length(asv_seqs)) 

# Fix the names in the rows and columns 
rownames(asv_rev_comp) <- asv_names
colnames(asv_rev_comp) <- asv_names

# Convert sequences to Biostrings
asv_dna <- DNAStringSet(asv_seqs) 

# Construct reverse complement
asv_rev_dna <- reverseComplement(asv_dna) 

# Now loop through every asv sequence to check 
for(i in 1:length(asv_seqs)){
  match_vec <- asv_dna[i] == asv_rev_dna
  asv_rev_comp[,i] <- match_vec
}

# Find how many TRUEs (matches) we have, divide by 2 because each pair occurs twice
cat("For", sum(asv_rev_comp) / 2, "ASVs, the reverse complement will need to be removed") 
```

<span style="color: red;">INTERPRETATION #4: Do you have any reverse complements in your dataset? : I did not detect any reverse complements in my dataset. The reverse complement check reported 0 ASVs that would need to be removed. This suggests that the ASVs are already in a consistent orientation and that reverse-complement artifacts are not inflating diversity in this dataset.</span> 

# 6. Sequencing Depth 

```{r seq-depth, fig.height=3, fig.width=8}
# The current data object
noControlsMock_ChlorosMitos_physeq

# Read metadata fresh from file
metadata_df <- read.csv("/local/workdir/glp56/Anglerfish_16S/data/metadata/SraRunTable.csv",
                        stringsAsFactors = FALSE)

# Build sequencing depth table and match metadata by Run accession
seqSums_df <- data.frame(
  names = sample_names(noControlsMock_ChlorosMitos_physeq),
  TotalSeqs = sample_sums(noControlsMock_ChlorosMitos_physeq),
  stringsAsFactors = FALSE
) %>%
  mutate(Run = sub("_R1_filtered\\.fastq\\.gz$", "", names)) %>%
  left_join(metadata_df, by = "Run")

# check
dim(seqSums_df)
head(seqSums_df)

# Show the depth of samples 
seqSums_df %>%
  dplyr::select(names, Run, TotalSeqs, isolation_source) %>%
  arrange(TotalSeqs) %>%
  head()

# Bar plot
numSeq_bar_plot <- 
  seqSums_df %>%
  arrange(TotalSeqs) %>%
  ggplot(aes(x = seq_along(names), y = TotalSeqs, fill = isolation_source)) + 
  geom_col(width = 0.9, color = NA) + 
  labs(y = "Read Depth", x = "Sample") + 
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.position = "none"
  )

# Histogram
numSeq_hist_plot <- 
  seqSums_df %>%
  ggplot(aes(x = TotalSeqs, fill = isolation_source)) + 
  geom_histogram(color = "black", bins = 20) + 
  labs(y = "# of Samples", x = "Read Depth") + 
  theme_bw() +
  theme(legend.position = "bottom",
        legend.title = element_blank())

# Density plot 
numSeq_density_plot <- 
  seqSums_df %>%
  ggplot(aes(TotalSeqs, fill = isolation_source, color = isolation_source)) +
  geom_density(alpha = 0.25, linewidth = 0.8) + 
  labs(x = "Read Depth", y = "Density") + 
  theme_bw() +
  theme(legend.position = "none")

# Put it all together 
numSeq_bar_plot + numSeq_hist_plot + numSeq_density_plot + 
  plot_annotation(tag_levels = "A")
```

<span style="color: red;">INTERPRETATION #5: What can you conclude about your sequencing depths? What are your sample distributions? Do you have "enough" reads? Are there any samples that should be removed at this step? : There is substantial variation in sequencing depth across samples, but most samples cluster at low to moderate read depths, with a smaller number of samples having much higher sequencing depth. The minimum sequencing depth was 153 reads, and the next few lowest-depth samples had 1,018, 4,038, 4,047, 4,283, and 5,332 reads. Most samples fall well above these lowest values, and the isolation source categories are all represented across the dataset. Overall, most samples appear to have enough reads for downstream analysis, but there are a few very low-depth samples. I think overall, I have enough reads after these steps.</span>

## Remove samples with few reads 

```{r rm-samps}
# What's the min seq depth? 
min(sample_sums(noControlsMock_ChlorosMitos_physeq))

# Keep all samples for now
preprocessed_physeq <- noControlsMock_ChlorosMitos_physeq

# What's the new min seq depth?
min(sample_sums(preprocessed_physeq))

# Final check of the sequencing depth for now
seqSums_df %>%
  dplyr::select(names, TotalSeqs) %>%
  arrange(TotalSeqs)
```

# 7. Add some metadata

***Likely, you do not need to do this step as you already have the data in your metadata! If so, you can skip this step.***

```{r add-metadata}
metadata_df <- 
  preprocessed_physeq %>% 
  sample_data() %>%
  data.frame()

if(!("names" %in% colnames(metadata_df))){
  metadata_df$names <- rownames(metadata_df)
}

num_seqs_df <- 
  preprocessed_physeq %>%
  sample_sums() %>%
  data.frame() %>%
  rownames_to_column(var = "names")

colnames(num_seqs_df)[2] <- "Sequencing_Depth"

metadata_df2 <- 
  metadata_df %>%
  left_join(num_seqs_df, by = "names")

rownames(metadata_df2) <- metadata_df2$names
sample_data(preprocessed_physeq) <- metadata_df2

colnames(data.frame(sample_data(preprocessed_physeq)))
```

# 8. Write out the data 

```{r save-physeq}
dir.create("data/03_PreProcessing", recursive = TRUE, showWarnings = FALSE)

save(preprocessed_physeq, file = "data/03_PreProcessing/preprocessed_physeq.RData")
```

# Final info for Reproducibility 

## Check Render Time
```{r stop-time}
# Take the time now that we are at the end of the script
end_time <- Sys.time()
end_time 

# Echo the elapsed time
elapsed_time <- round((end_time - start_time), 3)
elapsed_time
```

## Session Information

```{r session-info}
# Ensure reproducibility with package version information
devtools::session_info()
```