---
title: "06 Composition - Anglerfish"
author: "Griffin Patel"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

## Goals

In this file, we will perform compositional analysis of our scaled/noramlized/rarefied microbial dataset!

Load in scaled/normalized phyloseq object.
Calculate the relative abundances of taxonomic groups at various levels: A. Phylum B. Genus C. ASV
Plot them, and narrow in on specific taxnomic group of interest


## Set up the markdown with phyloseq file and colors

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  fig.path = "../figures/06_Composition/"
)

# Load in the packages 
pacman::p_load(tidyverse, devtools, DT, phyloseq, patchwork, 
               install = FALSE)

# Set up colors
tissue_colors <- c("Luminous" = "darkgoldenrod1",
                   "Non-luminous" = "deepskyblue")

isolation_shapes <- c("Caruncles" = 15,
                      "Esca" = 17,
                      "fin" = 15,
                      "gills" = 3,
                      "Guts" = 4,
                      "Illicium" = 8,
                      "skin" = 16)

# Load normalized phyloseq object
load("data/05_Ordination/scaled_physeq.RData")

# Reload metadata from the original sample sheet so the grouping variables are correct
metadata_df <- read.csv("data/metadata/SraRunTable.csv", stringsAsFactors = FALSE)

metadata_df$names <- paste0(metadata_df$Run, "_R1_filtered.fastq.gz")
metadata_df$tissue_group <- ifelse(metadata_df$isolation_source %in% c("Esca", "Caruncles"),
                                   "Luminous",
                                   "Non-luminous")

# Match metadata order to the phyloseq object
metadata_df <- metadata_df[match(sample_names(scaled_physeq), metadata_df$names), , drop = FALSE]
rownames(metadata_df) <- metadata_df$names

# Attach clean metadata
sample_data(scaled_physeq) <- sample_data(metadata_df)

# Look at the data 
scaled_physeq

# Intuition checks
min(sample_sums(scaled_physeq))
range(sample_sums(scaled_physeq))

# Confirm grouping variables are present and correct
table(data.frame(sample_data(scaled_physeq))$tissue_group, useNA = "ifany")
table(data.frame(sample_data(scaled_physeq))$isolation_source, useNA = "ifany")
```

## Taxonomic Analysis 

# A. Abundant Groups: Top 10 Phyla & ASVs

# A1: Top 10 phyla

```{r top-10-phyla}
# Create a phylum-level dataframe
phylum_df <- 
  scaled_physeq %>%
  tax_glom(taxrank = "Phylum") %>%
  transform_sample_counts(function(x) {x / sum(x)}) %>%
  psmelt()

## What are the phylum abundances?
phylum_df %>%
  group_by(Phylum) %>%
  summarize(mean_PercAbund = round(mean(Abundance), digits = 4), .groups = "drop") %>%
  arrange(-mean_PercAbund) %>%
  datatable()

# Make a list of the top 10 phyla
top10_phyla <- 
  phylum_df %>%
  group_by(Phylum) %>%
  summarize(mean_PercAbund = mean(Abundance), .groups = "drop") %>%
  arrange(-mean_PercAbund) %>%
  head(n = 10) %>%
  pull(Phylum)

# Plot the top 10 phyla with tissue_group on the x-axis
phylum_df %>%
  dplyr::filter(Phylum %in% top10_phyla) %>%
  ggplot(aes(x = tissue_group, y = Abundance, fill = Phylum, color = Phylum)) + 
  facet_wrap(Phylum ~ ., scales = "free_y", nrow = 2) + 
  geom_jitter(width = 0.15, alpha = 0.7) +
  geom_boxplot(outlier.shape = NA, alpha = 0.4) + 
  theme_bw() + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1),
        axis.title.x = element_blank())
```

# A2: Top 15 ASVs

```{r top-15-ASVs}
# Calculate ASV relative abundance
ASV_df <- 
  scaled_physeq %>%
  prune_taxa(taxa_sums(.) >= 211, .) %>%
  transform_sample_counts(function(x) {x / sum(x)}) %>%
  psmelt() %>%
  dplyr::rename(ASV = OTU)

# Make a list of the top 15 ASVs in the dataset
top15_ASVs <- 
  ASV_df %>%
  group_by(ASV) %>%
  summarize(mean_PercAbund = mean(Abundance), .groups = "drop") %>%
  arrange(-mean_PercAbund) %>%
  head(n = 15) %>%
  pull(ASV)

# Show them
top15_ASVs

# Create colors for the phyla represented in the top 15 ASVs
top15_phyla <- 
  ASV_df %>%
  dplyr::filter(ASV %in% top15_ASVs) %>%
  dplyr::distinct(Phylum) %>%
  dplyr::pull(Phylum)

asv_phylum_colors <- c(
  "Pseudomonadota" = "royalblue3",
  "Campylobacterota" = "grey50",
  "Bacteroidota" = "darkolivegreen3"
)

# Keep only the phyla actually present in the top 15 ASV plot
asv_phylum_colors <- asv_phylum_colors[names(asv_phylum_colors) %in% top15_phyla]

# Plot the top 15 ASVs across tissue groups
ASV_df %>%
  dplyr::filter(ASV %in% top15_ASVs) %>%
  ggplot(aes(x = tissue_group, y = Abundance, fill = Phylum, color = Phylum)) + 
  facet_wrap(ASV ~ Genus, scales = "free_y", nrow = 3) + 
  geom_jitter(width = 0.15, alpha = 0.7) +
  geom_boxplot(outlier.shape = NA, alpha = 0.4) + 
  scale_fill_manual(values = asv_phylum_colors) + 
  scale_color_manual(values = asv_phylum_colors) + 
  theme_bw() + 
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1),
        axis.title.x = element_blank())
```

## Genus level, focusing on Enterovibrio 

```{r genus-level}
# Calculate genus relative abundance
genus_df <- 
  scaled_physeq %>%
  tax_glom(taxrank = "Genus") %>% 
  transform_sample_counts(function(x) {x / sum(x)}) %>%
  psmelt()

# Plot genus: categorical tissue groups
enterovibrio_genus_tissueGroup <- 
  genus_df %>%
  dplyr::filter(Phylum == "Pseudomonadota") %>%
  dplyr::filter(Genus %in% c("Enterovibrio")) %>%
  ggplot(aes(x = tissue_group, y = Abundance, 
             fill = tissue_group, color = tissue_group)) + 
  facet_wrap(. ~ Genus, scales = "free_y", nrow = 1) + 
  geom_boxplot(alpha = 0.5, outlier.shape = NA) + 
  geom_jitter() + 
  theme_bw() + 
  labs(title = "Enterovibrio Relative Abundance") + 
  scale_color_manual(values = tissue_colors) + 
  scale_fill_manual(values = tissue_colors) + 
  theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1),
        legend.position = "none")

# Show the plot
enterovibrio_genus_tissueGroup

# Plot genus: finer tissue categories
enterovibrio_genus_isolationSource <- 
  genus_df %>%
  dplyr::filter(Phylum == "Pseudomonadota") %>%
  dplyr::filter(Genus %in% c("Enterovibrio")) %>%
  ggplot(aes(x = isolation_source, y = Abundance,
             fill = tissue_group, color = tissue_group)) + 
  facet_wrap(. ~ Genus, scales = "free_y", nrow = 1) + 
  geom_boxplot(alpha = 0.5, outlier.shape = NA) + 
  geom_jitter() + 
  theme_bw() + 
  labs(title = "Enterovibrio by Tissue Type") + 
  scale_color_manual(values = tissue_colors) + 
  scale_fill_manual(values = tissue_colors) + 
  theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1),
        legend.position = "none")

# Collect the Enterovibrio plots
enterovibrio_genus_tissueGroup / enterovibrio_genus_isolationSource + 
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A")
```